---
title: "GBM Analysis"
output: html_document
date: "2022-10-26"
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library(patchwork)
library(naniar)
library(dplyr)
library(ggplot2)
library(forcats)
knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Importing data

```{r}
gbm_data = read_csv(
  'data/gbm_patient_data.csv') %>%
  janitor::clean_names() %>%
  mutate(
    age = 2022 - year_of_birth,
    ) %>%
  replace_with_na_all(condition = ~.x == 999) %>%
  replace_with_na_all(condition = ~.x == 9) 
skimr::skim(gbm_data)

```


## Exploratory data analysis
```{r}
ggplot(data = gbm_data, aes(x = gender), color = gender) + geom_bar()

ggplot(data = gbm_data, aes(x = age)) + geom_bar()

ggplot(data = gbm_data, aes(x = age_at_diagnosis)) + geom_bar()

ggplot(data = gbm_data, aes(x = line_of_therapy)) + geom_bar()

ggplot(data = gbm_data, aes(x = race)) + geom_bar()

ggplot(data = gbm_data, aes(x = patients_level_of_involvement_in_gbm)) + geom_bar()

ggplot(data = gbm_data, aes(x = patients_treatment_goals)) + geom_bar()

ggplot(data = gbm_data, aes(x = primary_insurance)) + geom_bar()

ggplot(data = gbm_data, aes(x = adequate_caretaker_support)) + geom_bar()

ggplot(data = gbm_data, aes(x = travel_time_to_your_office_30_min)) + geom_bar()

ggplot(data = gbm_data, aes(x = ecog_at_1st_line)) + geom_bar()

ggplot(data = gbm_data, aes(x = ecog_at_2nd_line)) + geom_bar()

ggplot(data = gbm_data, aes(x = mgmt_methylated)) + geom_bar()
       
ggplot(data = gbm_data, aes(x = egfr_mutated)) + geom_bar()

ggplot(data = gbm_data, aes(x = tp53_mutated)) + geom_bar()

ggplot(data = gbm_data, aes(x = idh1_idh2_mutated)) + geom_bar()

ggplot(data = gbm_data, aes(x = pd_l1_overexpressed)) + geom_bar()

ggplot(data = gbm_data, aes(x = percent_of_tumor_mass_surgically_resected)) + geom_bar()

ggplot(data = gbm_data, aes(x = regimen_in_1st_line)) + geom_bar()

ggplot(data = gbm_data, aes(x = regimen_in_1st_line_other)) + geom_bar()

ggplot(data = gbm_data, aes(x = regimen_in_2nd_line)) + geom_bar()

ggplot(data = gbm_data, aes(x = regimen_in_2nd_line_other)) + geom_bar()

```



## Comorbidity analysis

Tidying dataframe:
```{r}
comorbidity_df = read_csv(
  'data/gbm_patient_data.csv') %>%
  janitor::clean_names() %>%
  mutate(
    age = 2022 - year_of_birth,
    ) %>%
  replace_with_na_all(condition = ~.x == 999) %>%
  replace_with_na_all(condition = ~.x == 9) %>%
  select(comorbidity_other_specify_29, everything()) %>%
  pivot_longer(
    comorbidity_renal_impairment:comorbidity_none,
    names_to = "comorbidity_type",
    names_prefix = "comorbidity_",
    values_to = "comorbidity",
  ) %>%
  select(patient_id, comorbidity_type, comorbidity)
```


Table of number of comorbidities per patient
```{r}
comorbidity_df %>%
  group_by(patient_id) %>%
  filter(comorbidity_type != "none") %>%
  filter(comorbidity == 1) %>%
  print(n = 850) %>%
  summarise(n_obs = n()) %>%
  knitr::kable()
```

Average number of comorbidities per patient 
```{r}
comorbidity_df %>%
  group_by(patient_id) %>%
  filter(comorbidity_type != "none") %>%
  filter(comorbidity == 1) %>%
  mutate(n_obs = n()) %>%
  select(patient_id, n_obs) %>%
  distinct %>%
  ungroup(patient_id) %>%
  summarise(
            mean = mean(n_obs),
            sd = sd(n_obs)) %>%
  knitr::kable()
```

distribution of number of comorbidities 
```{r}
comorbidity_df %>%
  group_by(patient_id) %>%
  filter(comorbidity_type != "none") %>%
  filter(comorbidity == 1) %>%
  mutate(n_obs = n()) %>%
  select(patient_id, n_obs) %>%
  distinct %>%
  ggplot(aes(x = n_obs)) + 
  geom_bar(fill = "lightblue") +
  labs(
    title = "Distribution of Number of Comorbidities per Patient",
    x = "Number of Comorbidities")
```


distribution of comorbidities
```{r}
comorbidity_df %>%
  filter(comorbidity_type != "none") %>%
  filter(comorbidity == 1) %>%
  group_by(comorbidity_type) %>%
  summarise(count = n()) %>%
  mutate(comorbidity_type = fct_reorder(comorbidity_type, count)) %>%
  ggplot(aes(x = comorbidity_type, y = count, color = comorbidity_type, fill = comorbidity_type)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Distribution of Comorbidity Types among Patients",
    x = "Comorbidities Types") +
    theme(legend.position = "bottom")
```

comorbidity - other specified:
```{r}
gbm_data %>%
  drop_na(comorbidity_other_specify_29) %>%
  ggplot(aes(x = comorbidity_other_specify_29)) + geom_bar()
```
8 patients specified - specific comorbidities include: 
aortic aneurism, arthritis, bipolar, disc disease, elevated cholesterol, herniated disc surgery, obese, parkinsons

## mutation data analysis 
