---
title: "GBM Analysis"
output: html_document
date: "2022-10-26"
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library(patchwork)
library(naniar)
library(dplyr)
library(ggplot2)
library(forcats)
knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Importing data

```{r}
gbm_data = read_csv(
  'data/gbm_patient_data.csv') %>%
  janitor::clean_names() %>%
  mutate(
    age = 2022 - year_of_birth,
    ) %>%
  replace_with_na_all(condition = ~.x == 999) %>%
  replace_with_na_all(condition = ~.x == 9) 
skimr::skim(gbm_data)

```


## Exploratory data analysis
```{r, include = FALSE}
ggplot(data = gbm_data, aes(x = gender), color = gender) + geom_bar()

ggplot(data = gbm_data, aes(x = age)) + geom_bar()

ggplot(data = gbm_data, aes(x = age_at_diagnosis)) + geom_bar()

ggplot(data = gbm_data, aes(x = line_of_therapy)) + geom_bar()

ggplot(data = gbm_data, aes(x = race)) + geom_bar()

ggplot(data = gbm_data, aes(x = patients_level_of_involvement_in_gbm)) + geom_bar()

ggplot(data = gbm_data, aes(x = patients_treatment_goals)) + geom_bar()

ggplot(data = gbm_data, aes(x = primary_insurance)) + geom_bar()

ggplot(data = gbm_data, aes(x = adequate_caretaker_support)) + geom_bar()

ggplot(data = gbm_data, aes(x = travel_time_to_your_office_30_min)) + geom_bar()

ggplot(data = gbm_data, aes(x = ecog_at_1st_line)) + geom_bar()

ggplot(data = gbm_data, aes(x = ecog_at_2nd_line)) + geom_bar()

ggplot(data = gbm_data, aes(x = mgmt_methylated)) + geom_bar()
       
ggplot(data = gbm_data, aes(x = egfr_mutated)) + geom_bar()

ggplot(data = gbm_data, aes(x = tp53_mutated)) + geom_bar()

ggplot(data = gbm_data, aes(x = idh1_idh2_mutated)) + geom_bar()

ggplot(data = gbm_data, aes(x = pd_l1_overexpressed)) + geom_bar()

ggplot(data = gbm_data, aes(x = percent_of_tumor_mass_surgically_resected)) + geom_bar()

ggplot(data = gbm_data, aes(x = regimen_in_1st_line)) + geom_bar()

ggplot(data = gbm_data, aes(x = regimen_in_1st_line_other)) + geom_bar()

ggplot(data = gbm_data, aes(x = regimen_in_2nd_line)) + geom_bar()

ggplot(data = gbm_data, aes(x = regimen_in_2nd_line_other)) + geom_bar()

```



## Comorbidity analysis

Tidying dataframe:
```{r}
comorbidity_df = read_csv(
  'data/gbm_patient_data.csv') %>%
  janitor::clean_names() %>%
  mutate(
    age = 2022 - year_of_birth,
    ) %>%
  replace_with_na_all(condition = ~.x == 999) %>%
  replace_with_na_all(condition = ~.x == 9) %>%
  select(comorbidity_other_specify_29, everything()) %>%
  pivot_longer(
    comorbidity_renal_impairment:comorbidity_none,
    names_to = "comorbidity_type",
    names_prefix = "comorbidity_",
    values_to = "comorbidity",
  ) %>%
  select(patient_id, comorbidity_type, comorbidity)
```


Table of number of comorbidities per patient
```{r}
comorbidity_df %>%
  group_by(patient_id) %>%
  filter(comorbidity_type != "none") %>%
  filter(comorbidity == 1) %>%
  print(n = 850) %>%
  summarise(n_obs = n()) %>%
  knitr::kable()
```

Average number of comorbidities per patient 
```{r}
comorbidity_df %>%
  group_by(patient_id) %>%
  filter(comorbidity_type != "none") %>%
  filter(comorbidity == 1) %>%
  mutate(n_obs = n()) %>%
  select(patient_id, n_obs) %>%
  distinct %>%
  ungroup(patient_id) %>%
  summarise(
            mean = mean(n_obs),
            sd = sd(n_obs)) %>%
  knitr::kable()
```

distribution of number of comorbidities 
```{r}
comorbidity_df %>%
  group_by(patient_id) %>%
  filter(comorbidity_type != "none") %>%
  filter(comorbidity == 1) %>%
  mutate(n_obs = n()) %>%
  select(patient_id, n_obs) %>%
  distinct %>%
  ggplot(aes(x = n_obs)) + 
  geom_bar(fill = "lightblue") +
  labs(
    title = "Distribution of Number of Comorbidities per Patient",
    x = "Number of Comorbidities")
```


distribution of comorbidities
```{r}
comorbidity_df %>%
  filter(comorbidity_type != "none") %>%
  filter(comorbidity == 1) %>%
  group_by(comorbidity_type) %>%
  summarise(count = n()) %>%
  mutate(comorbidity_type = fct_reorder(comorbidity_type, count)) %>%
  ggplot(aes(x = comorbidity_type, y = count, color = comorbidity_type, fill = comorbidity_type)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Distribution of Comorbidity Types among Patients",
    x = "Comorbidities Types") +
    theme(legend.position = "bottom")
```

comorbidity - other specified:
```{r, include = FALSE}
gbm_data %>%
  drop_na(comorbidity_other_specify_29) %>%
  ggplot(aes(x = comorbidity_other_specify_29)) + geom_bar()
```
8 patients specified - specific comorbidities include: 
aortic aneurism, arthritis, bipolar, disc disease, elevated cholesterol, herniated disc surgery, obese, parkinsons

## ecog comparison 
```{r}
ecog_data =
  gbm_data %>%
  select(patient_id, ecog_at_1st_line, ecog_at_2nd_line)

ggplot(data = ecog_data, aes(x = ecog_at_1st_line)) + 
  geom_bar(fill = "lightblue") +
  labs(
    title = "ECOG Distribution at 1st Line",
    x = "ECOG Performance Status Scale") +
    theme(legend.position = "bottom") +

ggplot(data = ecog_data, aes(x = ecog_at_2nd_line)) + 
  geom_bar(fill = "lightblue") +
  labs(
    title = "ECOG Distribution at 2nd Line",
    x = "ECOG Performance Status Scale") +
    theme(legend.position = "bottom")
```

##Regimen 1st vs 2nd
```{r}
gbm_data %>%
  count(regimen_in_1st_line) %>%
  mutate(regimen_in_1st_line = fct_reorder(regimen_in_1st_line, n)) %>%
  ggplot(aes(x = regimen_in_1st_line, y = n)) + geom_bar(stat="identity", fill = "lightblue") + coord_flip() +
  

gbm_data %>%
  count(regimen_in_2nd_line) %>%
  mutate(regimen_in_2nd_line = fct_reorder(regimen_in_2nd_line, n)) %>%
  ggplot(aes(x = regimen_in_2nd_line, y = n)) + geom_bar(stat="identity", fill = "lightblue") + coord_flip()


```


1st line ecog by treatment and 2nd line ecog by treatment
```{r}
ecog_data =
  gbm_data %>%
  select(patient_id, ecog_at_1st_line, ecog_at_2nd_line, regimen_in_1st_line, regimen_in_2nd_line)

ggplot(data = ecog_data, aes(x = ecog_at_1st_line, fill = regimen_in_1st_line)) + 
  geom_bar() +
  labs(
    title = "ECOG Distribution by Regimen at 1st Line Treatment",
    x = "ECOG at 1st Line Treatment") +
    theme(legend.position = "bottom") +
  
  ggplot(data = ecog_data, aes(x = ecog_at_2nd_line, fill = regimen_in_2nd_line)) + 
  geom_bar() +
  labs(
    title = "ECOG Distribution by Regimen at 2nd Line Treatment",
    x = "ECOG at 2nd Line Treatment") +
    theme(legend.position = "bottom")
```




## mutation analysis
tidying data
```{r}
mutation_df =
  gbm_data %>%
  select(patient_id, gender, mgmt_methylated:pd_l1_overexpressed) %>%
  pivot_longer(
    mgmt_methylated:pd_l1_overexpressed,
    names_to = "mutation_type",
    values_to = "mutation"
  )
```

table of number of mutations per patient
```{r}
mutation_df %>%
  group_by(patient_id) %>%
  filter(mutation == 1) %>%
  summarise(n_obs = n()) %>%
  knitr::kable()
```

average number of mutations per patient
```{r}
mutation_df %>%
  group_by(patient_id) %>%
  filter(mutation == 1) %>%
  mutate(n_obs = n()) %>%
  select(patient_id, n_obs) %>%
  distinct %>%
  ungroup(patient_id) %>%
  summarise(
            mean = mean(n_obs),
            sd = sd(n_obs)) %>%
  knitr::kable()
```

distribution of number of mutations
```{r}
mutation_df %>%
  group_by(patient_id) %>%
  filter(mutation == 1) %>%
  mutate(n_obs = n()) %>%
  select(patient_id, n_obs) %>%
  distinct %>%
  ggplot(aes(x = n_obs)) + 
  geom_bar(fill = "lightblue") +
  labs(
    title = "Distribution of Number of Mutations per Patient",
    x = "Number of Mutations")
```

distribution of mutations
```{r}
mutation_df %>%
  filter(mutation == 1) %>%
  group_by(mutation_type) %>%
  summarise(count = n()) %>%
  mutate(mutation_type = fct_reorder(mutation_type, count)) %>%
  ggplot(aes(x = mutation_type, y = count, color = mutation_type, fill = mutation_type)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Distribution of Mutation Types among Patients",
    x = "Mutation Types") +
    theme(legend.position = "bottom")
```

gender distribution across mutations
```{r}
mutation_gender_df =
  mutation_df %>%
    filter(mutation == 1) %>%
    group_by(patient_id) %>%
    mutate(count = n()) %>%
    mutate(mutation_type = fct_reorder(mutation_type, count))

ggplot(data = mutation_gender_df, (aes(x = reorder(mutation_type, -count), fill = gender))) + 
  geom_bar() +
  coord_flip() +
  labs(
    title = "Gender Distribution of Mutation Types among Patients",
    x = "Mutation Types") +
    theme(legend.position = "bottom")
```

## demographic plots
```{r}
ggplot(data = gbm_data, aes(x = gender), color = gender) + geom_bar(fill = "lightblue") +
    labs(
    title = "Gender Distribution of Patients",
    x = "Gender") +
    theme(legend.position = "bottom")

ggplot(data = gbm_data, aes(x = age_at_diagnosis)) + geom_bar(fill="lightblue") +
    labs(
    title = "Age of Diagnosis Distribution of Patients",
    x = "Age") +
    theme(legend.position = "bottom")

gbm_data %>%
  mutate(line_of_therapy = as.character(line_of_therapy)) %>%
  ggplot(aes(x = line_of_therapy)) + geom_bar(fill="lightblue") +
    labs(
    title = "Line of Therapy Distribution of Patients",
    x = "Line of Therapy") +
    theme(legend.position = "bottom")

gbm_data %>%
  count(race) %>%
  mutate(race = fct_reorder(race, -n)) %>%
  ggplot(aes(x = race, y = n)) + geom_bar(fill="lightblue", stat = "identity") +
    labs(
    title = "Race Distribution of Patients",
    x = "Race") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

gbm_data %>%
  count(primary_insurance) %>%
  mutate(primary_insurance = fct_reorder(primary_insurance, -n)) %>%
  ggplot(aes(x = primary_insurance, y = n)) + geom_bar(fill = "lightblue", stat = "identity") +
    labs(
    title = "Primary Health Insurance Distribution of Patients",
    x = "Primar Health Insurance") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```


